---
title: "Bobo the amoeba"
output: html_document
---

## Problem statement
*Bobo the amoeba has a 25%, 25%, and 50% chance of producing 0, 1, or 2 offspring, respectively. Each of Bobo’s descendants also have the same probabilities. What is the probability that Bobo’s lineage dies out?*

## Analytical solution

Let $X$ be a random variable describe the event "the population eventually dies out". We are asked to find $P(X)$.
For $i$ in ${0,1,2}$ let $A_i$ be a r.v. describing the event that Bobo turns into $i$ amoebas. It follows that

 1. $P(X|A_0) = 1$; Bobo dies without an offspring
 2. $P(X|A_1) = P(X)$; Bobo has one offspring, which will retain its reproduction probabilities
 3. $P(A|X_2) = P(X)^2$; Bobo has two offspring. Both will retain its reproduction probabilities

By the law of total probability we can now write $P(X)$ as 
$$P(X) = \frac{1}{4} P(X|A_0) + \frac{1}{4} P(X|A_1) + \frac{1}{2} P(X|A_2)$$

by doing simple algebra we reduce the equation to
$0 = \frac{1}{4}  + P(X) (-\frac{3}{4}  + \frac{1}{2} P(X))$

which has roots $P(X) = \frac{1}{2}$ and $P(X) = 1$.

Which one should we pick? Intuitively we can expect P(X) to be < 1, but let's validate this by means of simulation.

## Simulation
We can simulate a few iterations of Bobo's reproductive process and count how many times its lineage died out. If we do this for a large enough number of iterations, we should get an asymptotic value good enough to discriminate between the two roots. It's also a good excuse to brush up some R.

```{r}
library('ggplot2')
```

We can encode the probabilities $P(A_i)$ as 
```{r}
p <- c(0.25, 0.25, 0.50)
```

To generate an offspring wrt the given proabilities *p* we use a function that simulates a multinomial distribution.
```{r}
offspring <- function(v) {
  r <- runif(n=1)
  inP <- -1
  for (k in 1:length(p)) {
    inP <- k
    if ((k == 1) & (r < v[k])) {
      inP <- k-1
      break
    }
    else if ( (k>1) & (r < sum(v[1:k])) ) {
      inP <- k-1
      break
    }
  }
  return(inP)
}
```

We can refactor this code in a more performant, vectorized, form as:
```{r}
offspring <- function(v) {
  match(1, rmultinom(1, 1, p))-1
}
```

To convince ourselves that the multinomial simulation works, let's plot the frequency of each offspring over a few thousand iterations.
```{r}
t <- replicate(10000, offspring(p))

df <- as.data.frame(t)
colnames(df) <- c("outcomes")
ggplot(data=df, aes(x=as.factor(outcomes))) + 
  geom_histogram() +
  xlab("outcomes") + 
  ggtitle("Number of offspring")
```

### Simulation runs
```{r}
max_iterations <- 1000
max_generations <- 100
```
Finally let's simulate Bobo offspring over time. How many runs resulted in the offspring dying off within `{r} max_generations` out of `{r} max_iterations`?
```{r}
is_extinct <-function(generations) {
    offs <- 1 # We start with only Bobo
    next_offs <- 0 # offspring of the next generation
    
    while (offs <= generations) { # loop through generations  
      if (offs == 0) { # lineage died out
        break
      }
      next_offs <- 0 # offspring of the current generation
      for (j in 1:offs) {
        next_offs <- next_offs + offspring(p)
      }
      offs <- next_offs
    }
    return(as.integer(offs < generations)) # TRUE if the lineage dies out
}

# count the number of times Bobo's lineage died off. Do it for 10 runs of
# max_iterations each
res <- replicate(10, sum(replicate(max_iterations, is_extinct(max_generations))) / max_iterations)
```

Results suggest that Bobo's lineage will go extinct with $P(X) = \frac{1}{2}$.
```{r}
summary(res)
```

## Analytical solution, reprise
More formally, this problem can be modelled as a branching process; that is, we can look at it as a Markov chain where the state $A_n$ represents the number of amoebas at generation (or time) $n$, and transition probabilities can be written in the from $P(A_{n+1} = i |A_n =k) = P W_1 + W_2 + ... + W_k = i)$, where $W_1, ..., W_k$ are independent random variables each representing the offspring distribution. This leads to a recursive representation of the problem, where each $W_k$ models a branching process. We observe that if $A_n$ reaches 0, it says there (0 is an absorbing state). It is known (theorem) that when an individual has less than one offspring on average, denoted as $\mu \leq 1$, its lineage will die out with $P("extinction") = 1$. When $\mu \gt 1$ there are two solutions for the recursive sequence $P(X) = \frac{1}{4} P(X|A_0) + \frac{1}{4} P(X|A_1) + \frac{1}{2} P(X|A_2)$ and the extinction probability is the smallest of the two.
Analysis and proofs can be found at [this link](https://www.math.ucdavis.edu/~gravner/MAT135B/materials/ch14.pdf)

In our case the offspring expectation is $\mu = 1.25$; it follows that Bobo's lineage will go extinct with probability $P(X) = \frac{1}{2}$.
